Evaluating infectious disease forecasts \\ with allocation scoring rules

### 2024-01-29 alloscore_manuscript/alloscore-application-shorter.Rnw
compromise way to make percap

Infectious disease forecasting models have emerged as important tools in public health. Predictions of disease dynamics have been used to inform decision making about a wide variety of measures to reduce disease spread and/or mitigate the severity of disease outcomes. For example, estimates of expected onset of flu season have been used to aid national vaccination strategies \citep{igboh2023timing}, and forecasts of Ebola and diptheria dynamics have been used to allocate resources \citep{meltzer2014estimating, rainisch2015regional,camacho2015-ebola-bed,finger_real-time_2019}. \cite{bertsimas2021predictionsCOVID} developed tools to guide decision making from infectious disease forecasts, which have been used to inform allocation of limited medical supplies such as ventilators, ICU capacity planning, and vaccine distribution strategy. Models developed by \cite{fox_real-time_2022} have been used to inform resource and care site planning, as well as community guidelines for masking, traveling, dining and shopping (University of Texas, 2022)\nocite{utnews2022}.
%In April of 2022, the Centers for Disease Control and Prevention (CDC) announced the launch of the Center for Forecasting and Outbreak Analytics (CFA) to translate disease forecasts into decision making (CDC, 2022)\nocite{cdc2022cfa}, indicating that this has been identified as an important direction at the highest levels of government public health response.
%The value of infectious disease forecasts has typically been measured by how closely they predict disease outcomes such as cases, hospitalizations or deaths using, for example, root mean square error (RMSE) \citep{papastefanopoulos2020covid} or weighted interval score (WIS) \citep{bracher2021evaluating}.  However, recently authors have been calling for evaluating forecasts through their impact on policy \citep{marshall2023predictions, bilinski_adaptive_2023}.

@@

In section \ref{sec:application}, we present an illustrative analysis using the allocation score to evaluate forecasts of hospital admissions in the US leading up to the Omicron wave in winter 2022.
This analysis is ``synthetic'' in that it does not correspond to an actual analysis that supported decision making in real time.
However, the framework described in this paper corresponds to real-world decisions that must be made by public health administrators around the globe, and could be adapted in the future for such real-time situations such as informing bed demand, ventilator usage, or doses needed for specific medications.
%For example, forecasts for districts in Sierra Leone of bed demand to care for patients with Ebola was the subject of a real-time modeling study in late 2014 and early 2015 \citep{camacho2015-ebola-bed}.
%And, in 2020, a model developed by an academic research group turned predictions of COVID-19 hospitalizations into estimates of ventilator usage and shortages. This framework was used by the Hartford HealthCare system in Connecticut ``to align ventilator supply with projected demand at a time where the [COVID-19] pandemic was on the rise'' \citep{bertsimas2021predictionsCOVID}.
%These examples illustrate the potential for forecasts to inform decisions about how to allocate limited supplies such as temporary hospital beds, ventilators, personal protective equipment, or other supplies that are known to be effective at reducing transmission or severity of disease.
However, we emphasize again that these studies did not take the step of evaluating forecasts based on the quality of the allocation decisions that they supported or could have been used to support.

@@

These scores illustrate a general result: allocation scores for a forecast will tend to be larger when the resource constraint is close to the observed need, because this is when it matters most which locations are allocated more or less resources. If the available resources are small compared to the observed need, any allocation of those limited resources will result in a large amount of unmet need. If the available resources are comparatively larger, it becomes less important which locations receive relatively more or fewer resources because all locations will receive enough resources to meet their need. In either of these extremes of resource availability, the avoidable unmet need that arises from the allocation suggested by a forecast (i.e., the forecast's allocation score) will tend to be small.

@@

The weighted interval score (WIS) was proposed in 2020 as a way to score forecasts that were being made in the early stages of the COVID-19 pandemic \citep{bracher2021evaluating}; equivalent scores had also been used in previous forecast evaluation efforts \cite[e.g.,][]{hong2016probabilisticEnergyForecasting}.
The WIS is a proper scoring rule for forecasts that use a set of quantiles to represent a probabilistic forecast distribution.
While pointing a reader interested in more mathematical detail to \cite{bracher2021evaluating}, we note simply that the WIS is a weighted sum of interval scores at different probability levels (e.g., 50\% prediction intervals, 80\% PIs, 95\% PIs, etc...).
Larger interval scores indicate less skillful forecasts.
An interval score consists of (a) the width of the interval, with larger intervals receiving higher scores (higher scores indicate less accuracy), and (b) a penalty if the interval does not cover the eventual observation, which increases the further away the interval is from the observed value.
Equivalently, the WIS can also be characterized as a weighted sum of quantile scores for each individual predictive quantile.
The quantile score for a particular quantile level assigns an asymmetric penalty to predictions that are too high or too low, with the relative sizes of the penalties set so that in expectation the score is minimized by the given quantile of the distribution.
The most commonly used version of WIS is one that uses an equal weighting of all quantile levels, in which case WIS approximates the continuous ranked probability score (CRPS), a commonly used score for probabilistic forecasts.
\emph{It is important to note that this weighting was proposed because the resulting score approximates the CRPS, and not because it aligned with any particular public health decision-making rationale.}

@@

<<score-computation>>=
tar_load("Kat15k_alloscores")
tar_load("score_data")
tar_load("truth_data")
tar_load("forecast_data")

@@

<<multi-loc-ranks, fig.width=8, fig.cap="Component-wise breakdowns of the allocation score (Panel A) and weighted interval score (Panel B), by location for forecasts of hospitalization admissions on January 3, 2022, for two selected models (JHUAPL-Bucky and CU-select). Panel A shows the observed resource need, in this case the observed number of hospitalizations, for each state, along with the hypothetical number of resources allocated to the given location based on the forecasts from each model. The number of available resources was fixed at 15,000 and forecasts from each model were used to determine an optimal allocation strategy before the resource need was known. For most locations the resource need exceeded the resources allocated, indicated by the `observed resource need' bar being larger than the `allocation' bar. Panel B shows the breakdown of the weighted interval score (WIS) into components of underprediction, overprediction and dispersion. Larger values of WIS indicate more error, and the full WIS score for each location can be decomposed into the three components shown here.">>=
## date points to target_end_date of 2022-01-10 to highlight
ref_date_for_plot <- "2021-12-20"
models_to_plot <- c("JHUAPL-Bucky", "CU-select")

@@

\section*{Acknowledgements}

We wish to thank the following individuals who contributed valuable comments and feedback on early versions of this work: Matthew Biggerstaff, Rebecca Borchering, Sebastian Funk, Melissa Kerr, and Jeffrey Shaman.

This work has been supported by the National Institutes of General Medical Sciences (R35GM119582) and the U.S. CDC(1U01IP001122). The content is solely the responsibility of the authors and does not necessarily represent the official views of NIGMS, the National Institutes of Health, or CDC.

>>>>

Infectious disease forecasting models have emerged as important tools in public health. Predictions of disease dynamics have been used to inform decision making about a wide variety of measures to reduce disease spread and/or mitigate the severity of disease outcomes. For example, estimates of expected onset of flu season have been used to aid national vaccination strategies \citep{igboh2023timing}, and forecasts of Ebola dynamics have been used to allocate surveillance resources \citep{meltzer2014estimating, rainisch2015regional}. \cite{bertsimas2021predictionsCOVID} developed tools to guide decision making from infectious disease forecasts, which have been used to inform allocation of limited medical supplies such as ventilators, ICU capacity planning, and vaccine distribution strategy. Models developed by \cite{fox_real-time_2022} have been used to inform resource and care site planning, as well as community guidelines for masking, traveling, dining and shopping (University of Texas, 2022)\nocite{utnews2022}.
%In April of 2022, the Centers for Disease Control and Prevention (CDC) announced the launch of the Center for Forecasting and Outbreak Analytics (CFA) to translate disease forecasts into decision making (CDC, 2022)\nocite{cdc2022cfa}, indicating that this has been identified as an important direction at the highest levels of government public health response.
%The value of infectious disease forecasts has typically been measured by how closely they predict disease outcomes such as cases, hospitalizations or deaths using, for example, root mean square error (RMSE) \citep{papastefanopoulos2020covid} or weighted interval score (WIS) \citep{bracher2021evaluating}.  However, recently authors have been calling for evaluating forecasts through their impact on policy \citep{marshall2023predictions, bilinski_adaptive_2023}.

@@

In section \ref{sec:application}, we present an illustrative analysis using the allocation score to evaluate forecasts of hospital admissions in the US leading up to the Omicron wave in winter 2022.
This analysis is ``synthetic'' in that it does not correspond to an actual analysis that supported decision making in real time.
However, the framework described in this paper corresponds to real-world decisions that must be made by public health administrators around the globe, and could be adapted in the future for such real-time situations.
For example, forecasts for districts in Sierra Leone of bed demand to care for patients with Ebola was the subject of a real-time modeling study in late 2014 and early 2015 \citep{camacho2015-ebola-bed}.
And, in 2020, a model developed by an academic research group turned predictions of COVID-19 hospitalizations into estimates of ventilator usage and shortages. This framework was used by the Hartford HealthCare system in Connecticut ``to align ventilator supply with projected demand at a time where the [COVID-19] pandemic was on the rise'' \citep{bertsimas2021predictionsCOVID}.
%These examples illustrate the potential for forecasts to inform decisions about how to allocate limited supplies such as temporary hospital beds, ventilators, personal protective equipment, or other supplies that are known to be effective at reducing transmission or severity of disease.
However, we emphasize again that these studies did not take the step of evaluating forecasts based on the quality of the allocation decisions that they supported or could have been used to support.

@@

These scores illustrate a general result: allocation scores for a forecast will tend to be larger when the resource constraint is close to the observed need, because this is when it matters most which locations are allocated more or less resources. If the resource constraint is very small, any allocation of those limited resources will result in a large amount of unmet need. If the resource constraint is very large, it becomes less important which locations receive relatively more or less resources because all locations will receive enough resources to meet their need. In either of these extremes of resource availability, the avoidable unmet need that arises from the allocation suggested by a forecast (i.e., the forecast's allocation score) will tend to be small.

@@

The weighted interval score (WIS) was proposed in 2020 as a way to score forecasts that were being made in the early stages of the COVID-19 pandemic \citep{bracher2021evaluating}; equivalent scores had also been used in previous forecast evaluation efforts \cite[e.g.,][]{hong2016probabilisticEnergyForecasting}.
The WIS is a proper scoring rule for forecasts that use a set of quantiles to represent a probabilistic forecast distribution.
While pointing a reader interested in more mathematical detail to \cite{bracher2021evaluating}, we note simply that the WIS is a weighted sum of interval scores at different probability levels (e.g., 50\% prediction intervals, 80\% PIs, 95\% PIs, etc...).
Larger interval scores indicate less skillful forecasts.
An interval score consists of (a) the width of the interval, with larger intervals receiving higher scores, and (b) a penalty if the interval does not cover the eventual observation, which increases the further away the interval is from the observed value.
Equivalently, the WIS can also be characterized as a weighted sum of quantile scores for each individual predictive quantile.
The quantile score for a particular quantile level assigns an asymmetric penalty to predictions that are too high or too low, with the relative sizes of the penalties set so that in expectation the score is minimized by the given quantile of the distribution.
The most commonly used version of WIS is one that uses an equal weighting of all quantile levels, in which case WIS approximates the continuous ranked probability score (CRPS), a commonly used score for probabilistic forecasts.
\emph{It is important to note that this weighting was proposed because the resulting score approximates the CRPS, and not because it aligned with any particular public health decision-making rationale.}

@@

<<score-computation>>=
tar_load("Kat15k_alloscores")
tar_load("score_data")
tar_load("truth_data")
tar_load("forecast_data")
tar_load("percap")

Kat15k_alloscores <- bind_rows(Kat15k_alloscores, percap |> filter(K == 15000))

@@

<<multi-loc-ranks, fig.width=8, fig.cap="Component-wise breakdowns of the allocation score (Panel A) and weighted interval score (Panel B), by location for forecasts of hospitalization admissions on January 3, 2022, for two selected models (JHUAPL-Bucky and CU-select). Panel A shows the observed resource need, in this case the observed number of hospitalizations, for each state, along with the hypothetical number of resources allocated to the given location based on the forecasts from each model. The number of available resources was fixed at 15,000 and forecasts from each model were used to determine an optimal allocation strategy before the resource need was known. For most locations the resource need exceeded the resources allocated, indicated by the `observed resource need' bar being larger than the `allocation' bar. Panel B shows the breakdown of the weighted interval score (WIS) into components of underprediction, overprediction and dispersion. Larger values of WIS indicate more error, and the full WIS score for each location can be decomposed into the three components shown here.">>=
## date points to target_end_date of 2022-01-10 to highlight
ref_date_for_plot <- "2021-12-20"
models_to_plot <- c("JHUAPL-Bucky", "CU-select", "percap")

@@

\section*{Acknowledgements}
This work has been supported by the National Institutes of General Medical Sciences (R35GM119582) and the U.S. CDC(1U01IP001122). The content is solely the responsibility of the authors and does not necessarily represent the official views of NIGMS, the National Institutes of Health, or CDC.

### 2024-01-29 alloscore_manuscript/alloscore-application-shorter.tex
compromise way to make percap

% latex table generated in R 4.3.1 by xtable 1.8-4 package
% Fri Dec 22 10:45:01 2023
\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
  \hline
Model & AS & MWIS & IAS centered at 15k & IAS uniform \\ 
  \hline
CU-select & 669 & 133 & 774 & 326 \\ 
  COVIDhub-ensemble & 873 & 159 & 1067 & 438 \\ 
  USC-SI\_kJalpha & 995 & 91 & 1216 & 1097 \\ 
  JHUAPL-Gecko & 1034 & 164 & 1141 & 418 \\ 
  MUNI-ARIMA & 1084 & 169 & 1248 & 440 \\ 
  COVIDhub-trained\_ensemble & 1089 & 169 & 1271 & 823 \\ 
  COVIDhub-baseline & 1175 & 170 & 1317 & 535 \\ 
  JHUAPL-Bucky & 1358 & 102 & 1566 & 1214 \\ 
  JHUAPL-SLPHospEns & 1540 & 129 & 1604 & 1102 \\ 
  UVA-Ensemble & 2469 & 213 & 2635 & 2494 \\ 
   \hline
\end{tabular}
\caption{Comparison of allocation scores (AS), mean weighted interval scores (MWIS), and two varieties of Integrated Allocation Scores (IAS). All metrics are shown for 10 models that made forecasts of hospital admissions for 2022-01-03. Results are sorted by AS. For all metrics, lower scores indicate better accuracy.} 
\label{tab:multi-k-scores}
\end{table}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/metrics-over-time-1} \caption[Hospital admissions and evaluation metrics over time]{Hospital admissions and evaluation metrics over time. Panel A shows the number of hospital admissions in the US as a whole due to COVID-19 on a sequence of 15 Mondays from December 2021 through March 2022. These are the dates for which forecasts were made and evaluated. A horizontal dashed line at 15,000 shows the assumed resource constraint $K$. Panel B shows allocation scores (AS) for each model's 14 day-ahead forecast, across all US states. The x-axis corresponds to the date for which the prediction was made. AS typically are high when the observed value is near to the constraint, which occurs during the last Monday in December (on the way up) and the last Monday in January (on the way down). Panel C shows the MWIS metric across weeks, averaged across all states. MWIS values tend to scale with the observed and predicted values, and the peak MWIS values happen around and just after the peak of the Omicron wave.}\label{fig:metrics-over-time}
\end{figure}

>>>>

% latex table generated in R 4.3.1 by xtable 1.8-4 package
% Mon Jan 29 15:31:56 2024
\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
  \hline
Model & AS & MWIS & IAS centered at 15k & IAS uniform \\ 
  \hline
CU-select & 669 & 133 & 774 & 326 \\ 
  COVIDhub-ensemble & 873 & 159 & 1067 & 438 \\ 
  USC-SI\_kJalpha & 995 & 91 & 1216 & 1097 \\ 
  JHUAPL-Gecko & 1034 & 164 & 1141 & 418 \\ 
  MUNI-ARIMA & 1084 & 169 & 1248 & 440 \\ 
  COVIDhub-trained\_ensemble & 1089 & 169 & 1271 & 823 \\ 
  COVIDhub-baseline & 1175 & 170 & 1317 & 535 \\ 
  JHUAPL-Bucky & 1358 & 102 & 1566 & 1214 \\ 
  JHUAPL-SLPHospEns & 1540 & 129 & 1604 & 1102 \\ 
  UVA-Ensemble & 2469 & 213 & 2635 & 2494 \\ 
   \hline
\end{tabular}
\caption{Comparison of allocation scores (AS), mean weighted interval scores (MWIS), and two varieties of Integrated Allocation Scores (IAS). All metrics are shown for 10 models that made forecasts of hospital admissions for 2022-01-03. Results are sorted by AS. For all metrics, lower scores indicate better accuracy.} 
\label{tab:multi-k-scores}
\end{table}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in `palette()`:\\\#\# ! Insufficient values in manual scale. 18 needed but only 17 provided.}}\end{kframe}
\end{knitrout}
